---
title: "STAA 566 HW4"
author: "Ryan Miller"
output: html_document
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---
TODO: UPDATE
\
The data for this assignment was retrieved from the [Gravitational Wave Open Science
Center](https://www.gw-openscience.org/). I chose this data because it is in
HDF5 format, which I've never used in R before so I had to learn something new (in
addition to the interactive plots!). The code will automatically download the data from
the website, unless a local file path is provided to it.

Working with the data does require the use of the `hdf5r` [package](https://cran.r-project.org/web/packages/hdf5r/index.html), which can be
installed via the R terminal.

The data is of the first observed gravitational wave which took place in September
2015 and was produced by the merging of two black holes.
Data is shown from each of the two detectors `L1` (Livingston) and `H1` (Hanford). More
information can be found
[here](https://www.gw-openscience.org/events/GW150914/#:~:text=The%20event%20occurred%20at%20GPS,Hanford%20and%20LIGO%20Livingston%20observatories.).
This event is known as `GW150914`, which you will see in the plot titles and axes.

You will notice in the plots that the signal is currently indistinguishable from the
noise. The data needs to be processed in order for the signal to become clear, and it
turned out that I will need more time to implement this. My plan is to finish the signal
processing and include it in the shiny project. In that way the hope is that it will be
easy to compare the raw signal from the final product in an interactive way.

```{r, echo = FALSE, message = FALSE}
## Load required libraries
library(   hdf5r   )
library(  plotly   )
library(  pracma   )
library(    psd    )
library( tidyverse )
library(  viridis  )

## Clear function for ease (by default keeps all functions)
# Optional input E - character array of exceptions
# Optional input P - character array of inclusions
clear <- function( all = FALSE, E = "", P = "" )
{
    # By default don't clear the loaded data unless intentionally cleared
    if( ! all ){ E <- c( E, 'h1.5', 'l1.5', 'nrel' ) }

    # Get list of functions and exclude anything in E
    l <- setdiff( ls(envir=parent.frame()), c( lsf.str(envir=parent.frame()), E ) )

    # Add P to the list.
    if( all( P != "" ) ){ l <- c( l, P ) }

    # Remove the stuff
    rm( list = l, envir = parent.frame() )
}

## Clear things
clear( )
```
```{r, echo = FALSE, results = 'hide', message = FALSE}

#########################################################################################
## Get the data
#########################################################################################
## Caution: Downloading data takes may take some time. Recommend that you only do this
## once.
#########################################################################################
if( !exists( 'h1.5' ) | !exists( 'l1.5' )  | !exists( 'nrel' ) )
{
    ## Get strain data from H1 detector
    h1.file.loc  <- 'https://losc.ligo.org/s/events/GW150914/H-H1_LOSC_4_V1-1126259446-32.hdf5'
    h1.temp.file <- tempfile( fileext = '.hdf5' )
    download.file( h1.file.loc, h1.temp.file )
    h1.5    <- hdf5r::H5File$new( h1.temp.file, mode = 'r+' )
    file.remove( h1.temp.file );
    rm( list = c('h1.file.loc', 'h1.temp.file' ) )

    ## Get strain data from L1 detector
    l1.file.loc  <- 'https://losc.ligo.org/s/events/GW150914/L-L1_LOSC_4_V1-1126259446-32.hdf5'
    l1.temp.file <- tempfile( fileext = '.hdf5' )
    download.file( l1.file.loc, l1.temp.file )
    l1.5    <- hdf5r::H5File$new( l1.temp.file, mode = 'r+' )
    file.remove( l1.temp.file );
    rm( list = c('l1.file.loc', 'l1.temp.file' ) )

    ## Get numeric relativity template
    nrel <- read_table( 'https://losc.ligo.org/s/events/GW150914/GW150914_4_NR_waveform.txt',
                       col_names = FALSE )
}
```
```{r, echo = FALSE}

## Generate a time vector (in this data we know that H1 L1 are the same)
dt        <- h5attr( h1.5[['/strain/Strain']], 'Xspacing' )
gps.start <- h1.5[['/meta/GPSstart']][]
gps.end   <- gps.start + length( h1.5[['/quality/injections/Injmask']][] )
time      <- seq( from = gps.start, to = gps.end, by = dt )
dt2       <- time[2] - time[1]; if( dt != dt2 ){ dt = dt2 }

## Get the strain values for both detectors
strain.h1 <- h1.5[['/strain/Strain']][]
strain.l1 <- l1.5[['/strain/Strain']][]

```
```{r, echo = FALSE, fig.align = 'center'}

## Recreate HW2 (raw signal plot)
t.event <- 1126259462.422
t.off   <- 5
idx <- which( time >= t.event - t.off & time < t.event + t.off )

## Set up the plotly functions
pd.raw <- data.frame( L1 = strain.l1[ idx ], H1 = strain.h1[ idx ],
                     Time = time[ idx ] - t.event )
h.raw <- plot_ly( pd.raw, x = ~Time ) %>%
    add_lines( y = ~L1, name = 'L1' ) %>%
    add_lines( y = ~H1, name = 'H1' ) %>% rangeslider() %>% layout( hovermode = 'x' ) %>%
    layout( title = 'LIGO Raw Strain Near GW150914',
           xaxis = list( title = 'Time (s) since GW150914' ),
           yaxis = list( title = 'Strain' ) )

h.raw
```
```{r, message = FALSE, echo = FALSE}
## Get spectral densities
fs     <- 4096 # Sampling frequency
psd.h1 <- psd::pspectrum( strain.h1, x.frqsamp = fs )
psd.l1 <- psd::pspectrum( strain.l1, x.frqsamp = fs )

## Whiten the signal
if( ! ( length( strain.h1 ) - length( strain.l1 ) ) )
{
    # Will error out after this if N is undefined; i.e., the lengths are different
    N    <- length( strain.h1 )
    EVEN <- !psd::mod( N, 2 )
}

if( EVEN )
# Even
{
    pos.freqs <- seq( from = 0, to = ( N / 2 - 1 ) / ( dt * N ), length.out = N/2)
    neg.freqs <- seq( from = -N/2 / (dt*N), to = -1 / (dt*N), length.out = N/2)
    #freqs     <- seq( from = 0, to = 1 / (2*dt), length.out = floor( N/2 ) + 1 )
}
# Odd
{
    pos.freqs <- seq( from = 0, to = ( (N-1)/2 / (dt*N) ), length.out = N/2 )
    neg.freqs <- seq( from = -((N-1)/2 / (dt*N)), to = -1 / (dt*N), length.out = N/2 )
    #freqs     <- seq( from = 0, to = ( (N-1)/2 ) / (dt*N), length.out = floor( N/2 ) + 1 )
}
freqs      <- c( pos.freqs, neg.freqs )

## Whiten H1
div        <-  sqrt( pracma::interp1( psd.h1$freq, psd.h1$spec, abs(freqs) ) / dt / 2 )
white.h1.f <- fft( strain.h1 ) / div
white.h1   <- Re( fft( white.h1.f, inverse = TRUE ) / N )

## Whiten L1
div        <-  sqrt( pracma::interp1( psd.l1$freq, psd.l1$spec, abs(freqs) ) / dt / 2 )
white.l1.f <- fft( strain.l1 ) / div
white.l1   <- Re( fft( white.l1.f, inverse = TRUE ) / N )

## Also have to do this for the relativistic model
# TODO: You are here.

# Shift by mean so they are both around zero, something. This is normally
# handled by the whitening process with Real-FFT's, which are not implemented in R.
white.h1 <- white.h1 - mean( white.h1 )
white.l1 <- white.l1 - mean( white.l1 )
# TODO: Also do for relativistic model

```
```{r, echo = FALSE}

## Plot the whitened signal
pd.white <- data.frame( L1 = white.l1[ idx ], H1 = white.h1[ idx ],
                     Time = time[ idx ] - t.event )
h.white <- plot_ly( pd.white, x = ~Time ) %>%
    add_lines( y = ~L1, name = 'L1' ) %>%
    add_lines( y = ~H1, name = 'H1' ) %>% rangeslider() %>% layout( hovermode = 'x' ) %>%
    layout( title = 'LIGO Whitened Strain Near GW150914',
           xaxis = list( title = 'Time (s) since GW150914' ),
           yaxis = list( title = 'Whitened Strain' ) )

h.white
```
```{r, echo = FALSE}

#########################################################################################
## Close all HDF5 stuff
#########################################################################################
#l1.5$close_all()
#h1.5$close_all()
```

## Code

```{r getlabs, results = 'hide', include = FALSE}
## Get all the labels
appLabs <- knitr::all_labels()
appLabs <- setdiff( appLabs, c('getlabs', appLabs[grep("^hide", appLabs)]) )
```

```{r show-code, ref.label = appLabs, echo = TRUE, eval = FALSE}
```

## References
A guide to LIGO-Virgo detector noise and extraction of transient gravitational-wave
signals; https://arxiv.org/pdf/1908.11170.pdf

Signal Processing with GW150914 Open Data;
https://www.gw-openscience.org/GW150914data/GW150914_tutorial.html
