---
title: "STAA 566 HW4"
author: "Ryan Miller"
output: html_document
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---
```{r, echo = FALSE, message = FALSE}
##########################################################################################
## Start Setup
##########################################################################################
## Load required libraries
library( gridExtra  )
library( tidyverse  )
library(  viridis   )
library( kableExtra )
library( gapminder  )
library( sparkline  )

## Clear function for ease (by default keeps all functions)
# Optional input E - character array of exceptions
# Optional input P - character array of inclusions
clear <- function( all = FALSE, E = "", P = "" )
{
    # By default don't clear the loaded data unless intentionally cleared
    if( ! all ){ E <- c( E, 'api.key' ) }

    # Get list of functions and exclude anything in E
    l <- setdiff( ls(envir=parent.frame()), c( lsf.str(envir=parent.frame()), E ) )

    # Add P to the list.
    if( all( P != "" ) ){ l <- c( l, P ) }

    # Remove the stuff
    rm( list = l, envir = parent.frame() )
}

## Clear things
clear( )

##########################################################################################
## End Setup
##########################################################################################
```
The following table presents life expectancy data for 142 countries from 1952 to 2007.

Each country has a trend plot, which includes the life expectancy from 1952 to 2007 in
five year increments. In addition, each country has the average life expectancy across the
entire period, the life expectancy in the first year, and the life expectancy in the last
year.

The trend plots include a tooltip so that when you hover, you get the year and life
expectancy value associated with it. I thought this was necessary to make the trend plots
worth while.

The last column includes a tool tip which is the percent change in life expectancy from
the first year (1952) to the last year (2007). I thought this was an easy way to get some
more useful information out of the table.

This table shows a lot of information in a very succinct way and allows you to
see trends across the time span as well as over all summary data. I find it interesting to
see the impacts of historical events.

The data comes from the `gapminder` package.

```{r, echo = FALSE, results = 'hide'}
## Calculate the average life expectancy over the time frame
avgLifeExp <- gapminder %>% group_by( country ) %>%
    summarize( avgLifeExp = mean( lifeExp ) )

## Some other useful information
years      <- seq( 1952, 2007, by = 5 )
#  [1] 1952 1957 1962 1967 1972 1977 1982 1987 1992 1997 2002 2007
nPeriod    <- length( years )
# [1] 12
years <- as.integer( years )
```
```{r, echo = FALSE}

## Create the trend plots
sparkline( 0 )
# Add a column
avgLifeExp <- avgLifeExp %>% mutate( plt_le = NA )
# Generate plots
for( i in levels( avgLifeExp$country ) )
{
    lifeExp <- gapminder %>% filter( country == i ) %>% pull( lifeExp )
    avgLifeExp[ which( avgLifeExp$country == i ), 'plt_le' ] <-
        spk_chr( lifeExp, xvalues = years,
                numberDigitGroupSep='', tooltipFormat = '{{x}}: {{y}}' )
}
```
```{r, echo = FALSE}

## Add the beginning and end year
DIGITS = 2
for( i in levels( avgLifeExp$country ) )
{
    fy <- unlist( gapminder %>% filter( year == min( years ) ) %>%
        dplyr::select( lifeExp ) )
    ly <- unlist( gapminder %>% filter( year == max( years ) ) %>%
        dplyr::select( lifeExp ) )

    p <- (ly - fy) / fy * 100
    avgLifeExp$firstYear <- fy
    avgLifeExp$lastYear  <- cell_spec( ly, tooltip = sprintf( '%0.2f%%', round(p, DIGITS)) )
}

## Order the table how I want
avgLifeExp <- avgLifeExp %>% dplyr::select( country, plt_le, avgLifeExp, firstYear,
                                           lastYear )
```
```{r, echo = FALSE}

## Generate the table
tableFull <- kbl( avgLifeExp, escape = FALSE,
    col.names = c('Country', 'Life Expectancy Trend', 'Avg. Life Expectancey',
                 sprintf('Life Exp. in %d', min(years)),
                 sprintf('Life Exp. in %d', max(years)) ),
    digits = DIGITS, align = 'c' ) %>%
kable_styling( fixed_thead = TRUE )

tableFull
```

## Code

```{r getlabs, results = 'hide', include = FALSE}
## Get all the labels
appLabs <- knitr::all_labels()
appLabs <- setdiff( appLabs, c('getlabs', appLabs[grep("^hide", appLabs)]) )
```

```{r show-code, ref.label = appLabs, echo = TRUE, eval = FALSE}
```
